---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Requiring packages

```{r echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(dplyr)
```


## Loading and preprocessing the data

Here we unzip, read the activity data and format time to analysis.

```{r echo=TRUE}
activityData <- read.csv(unz("activity.zip", "activity.csv"), sep = ",")
activityData$date <- as.Date(as.character(activityData$date), "%Y-%m-%d")
```


## What is mean total number of steps taken per day?

First we group data by date and sumarise the sum of steps
```{r echo=TRUE}
totalByDay <- group_by(activityData, date) %>% summarize(steps = sum(steps, na.rm = TRUE))
```

Result:
```{r echo=TRUE}
head(totalByDay, 10)
```

Now we get the mean and the median
```{r echo=TRUE}
meanSteps <- mean(totalByDay$steps, na.rm = TRUE)
medianSteps <- median(totalByDay$steps, na.rm = TRUE)
```

And add everything in a good informative plot
```{r echo=TRUE, fig.align='center'}
hist(totalByDay$steps, col = "turquoise3", xlab = "Total number of steps", main = "Total number of steps taken each day")

#Add lines in the mean and median
abline(v=meanSteps, col="red", lwd=2)
abline(v=medianSteps, col="black", lwd=2)

legend(
    "topright", # places a legend at the appropriate place
    c(paste0("Mean (", round(meanSteps,1)," steps)"),
      paste0("Median (", medianSteps," steps)")), # puts text in the legend
    pch = c("|","|"), # gives the legend appropriate symbols
    col=c("red", "black"),# gives the legend lines the correct color
    cex=0.8 # gives the legend box size
) 
```



## What is the average daily activity pattern?

First we need to group the data by interval and summarize the average.
```{r echo=TRUE, fig.align='center'}
groupInterval <- group_by(activityData, interval) %>%
            summarise(averageSteps = mean(steps, na.rm = TRUE))
```

Result:
```{r echo=TRUE, fig.align='center'}
head(groupInterval, 10)
```

Now we can get the interval with the maximum average number of steps.
```{r echo=TRUE}
maxSteps <- groupInterval[groupInterval$averageSteps == max(groupInterval$averageSteps), ]
```

Let's make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days.
```{r echo=TRUE, fig.align='center'}
with(groupInterval, plot(interval, averageSteps, type = "l", xlab = "5-minute interval", ylab = "Average number of steps"))
title(main="What is the average daily activity pattern?")

# Finally we can add a point to indicate the interval with the maximum number of steps
legendText <- paste0("\n \nMaximum number \nof steps (interval: ", maxSteps$interval,")")
points(maxSteps$interval, maxSteps$averageSteps, pch = 16, col = "red")
text(maxSteps$interval, maxSteps$averageSteps, labels = legendText, pos = 4)
```

## Imputing missing values

Note that there is some missing values in our data

```{r echo=TRUE}
# Number of lines with missing values
missingValuesCount <- nrow(activityData[is.na(activityData$steps), ])
print(missingValuesCount)
```

Has been there `r missingValuesCount` missing values, it can be changing the result, so let's
fill theses values using the average by interval data used in the previous question.

```{r echo=TRUE}
# Creating a new datset with original data
filledData <- activityData 

# Loop through the rows
for(i in 1:nrow(filledData)){
    # Verify if this row has missing value
    if(is.na(filledData[i, "steps"])){
        # Get row 5-minute interval
        rowInterval <- filledData[i, "interval"]
        
        # Filling in the missing value with the round of the mean value to this interval
        filledData[i, "steps"] <- round(groupInterval[groupInterval$interval == rowInterval, "averageSteps"], 0)
    }
}
```

In our new data set there is no more missing values

```{r echo=TRUE}
# Number of lines with missing values in new data set
nrow(filledData[is.na(filledData$steps), ])
```

And we can set the difference in the first rows
```{r echo=TRUE}
# Original data
head(activityData,10)

# Filled in data
head(filledData,10)
```

First we group data by date and sumarise the sum of steps
```{r echo=TRUE}
filledTotalByDay <- group_by(filledData, date) %>% summarize(steps = sum(steps, na.rm = TRUE))
```

Result:
```{r echo=TRUE}
head(filledTotalByDay, 10)
```

Now we get the mean and the median
```{r echo=TRUE}
meanSteps <- mean(filledTotalByDay$steps, na.rm = TRUE)
medianSteps <- median(filledTotalByDay$steps, na.rm = TRUE)
```

And add everything in a good informative plot
```{r echo=TRUE, fig.align='center'}
hist(filledTotalByDay$steps, col = "turquoise3", xlab = "Total number of steps", main = "Total number of steps taken each day")

#Add lines in the mean and median
abline(v=meanSteps, col="red", lwd=2)
abline(v=medianSteps, col="black", lwd=2, lty = 2)

legend(
    "topright", # places a legend at the appropriate place
    c(paste0("Mean (", round(meanSteps,1)," steps)"),
      paste0("Median (", medianSteps," steps)")), # puts text in the legend
    pch = c("|","|"), # gives the legend appropriate symbols
    col=c("red", "black"),# gives the legend lines the correct color
    cex=0.8 # gives the legend box size
) 
```

Without missing values the mean value increased and is closest to median

## Are there differences in activity patterns between weekdays and weekends?

It's a good idea to set locale category to show week days in English
```{r echo=TRUE}
Sys.setlocale("LC_TIME", "C")
unique(weekdays(filledData$date))
```

Now we will create a new factor variable to indicating whether a given date is a weekday or weekend day and group the data by interval and weekday
```{r echo=TRUE}
filledData$weekday <- "weekday"
filledData[weekdays(filledData$date) %in% c("Saturday", "Sunday"), "weekday"] <- "weekend"
filledData$weekday <- as.factor(filledData$weekday)

#Group by interval and weekday and summarize the average
groupIntervalWeekDay <- group_by(filledData, interval, weekday) %>%
    summarise(averageSteps = mean(steps, na.rm = TRUE))

head(groupIntervalWeekDay, 10)
```

We will use the ggplot2 library to make more easy to create panel plots
```{r echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(ggplot2)
```

```{r echo=TRUE, fig.align='center'}
qplot(interval, averageSteps, data = groupIntervalWeekDay,
      facets = weekday ~ ., geom = "line",
      xlab = "5-minute interval", ylab = "Number of steps",
      main = "Differences in activity \npatterns between weekdays and weekends")
```